---
title: "Danish Kings"
author: "Silas Bramsen, Olivia Clara Munk-Olsen, Freja Bille Deichmann og Laura Holm Løjborg
"
date: "12.5.25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator:_______  

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs
library(tidyverse)
kings1 <- read.csv("data/kongerække.csv")
head(kings1)
glimpse(kings1)
class(kings1)

kings2 <- read_csv("data/kongerække.csv")
head(kings2)
glimpse(kings2)
class(kings2)

kings3 <- read.csv2("data/kongerække.csv")
head(kings3)
glimpse(kings3)
class(kings3)

kings4 <- read_csv2( "data/kongerække.csv", na="NA")
head(kings4)
glimpse(kings4)
class(kings4)



```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object
  The one that is a tidyverse function is the one that R reads as a   tibble, “tbl”. In this case, with this particular dataset, the ones that are tidyverse functions are the ones that include underscore "_" rather than a full stop. So kings2 and kings4 are the ones read as a tibble.  
  

  
2. What is the result of running `class()` on the `kings` object created with a tidyverse function.
  When running 'class(kings)' on kings2 and kings4 the result is: "spec_tbl_df"   "tbl_df"      "tbl"         "data.frame" 

  
3. How many columns does the object have when created with these different functions?
  By using the ncol() function the obejct (kings <- kings4), is revealed to have 8 columns 

  
  
4. Show the dataset so that we can see how R interprets each column

```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- kings4

class(kings)

head(kings)

tail(kings)

```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`

  The dataset contains missing fata, but by using the function na="NA", when loading the csv.file it is already omitted 



Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.
  
  We start by selecting the names, start and end of reign, followed by pipe. Then we mutate start and end of reign into a new column called "duration"


```{r}
# YOUR CODE

kings_duration <- select(kings,Navn,Regering_Start,Regering_Slut) %>% 
  mutate(duration=Regering_Slut-Regering_Start)
  


```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`

  Using the kings_duration object we created before, we select the duration and summarise it, and calling it ´genm´.



```{r}
# YOUR CODE
kings_duration %>% 
  select(duration) %>% 
  summarise(genm=mean(duration, na.rm = TRUE))



```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`
  
  We chose to name the object for rulers that ruled longer than average ´over_genm´ 
 

```{r}
# YOUR CODE

over_genm <- kings_duration %>% 
  filter(duration>19.92727)
count(over_genm)
over_genm


```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# YOUR CODE

top3 <- over_genm %>%
  arrange(desc(duration)) %>%
  slice(1:3) %>%
  mutate(days = duration * 365.25)

top3 %>% select(Navn, days)


  
```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# YOUR CODE
```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


